Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-09 submission/manuscript.Rmd
egg sperm genes

- Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes

@@

Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation). The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F). The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G). Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.

@@

While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions. It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.

To comprehensively assess the impact of gonadal sex on germline gene misexpression, we filtered our list of germline-enriched genes for egg (XX) or sperm (XY) biased genes. We determined male and female biased expression  using RNAseq data from wild-type male and female mice at embryonic day 12, 14, and 16. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased 
germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 2A).

>>>>

@@

Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation). The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F). The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G). Thus, while female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, it is likely independent of potential XCI defects.

@@

While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess if germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L). _Kdm5c_ mutant male and female-specific germline DEGs were also not biased to the corresponding germ cell sex, indicating differences between male and female germline gene dysregulation not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate sex influences the degree of gene misexpression with loss of KDM5C, but not the sex of germ cell-enriched genes.
